---
title: "HR.Function"
author: "Kosar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Load required libraries
```{r}
library(survival)
library(dplyr)
library(purrr)
library(tibble)
library(stats)
library(openxlsx)
library(ggplot2)
library(cowplot)
library(grid)
```
# Section 1: Helper function for HR
```{r}
analyze_pathways_survival <- function(df, pathway_cols, tag = NULL, save_results = FALSE, output_dir = ".") {
  
  endpoint_definitions <- list(
    OS  = list(time = "OS(Months)",   event = "OS status"),
    DSS = list(time = "OS(Months)",   event = "DSS"),
    RFS = list(time = "RFS (Months)", event = "RFS"),
    MFS = list(time = "OS(Months)",   event = "MFS")
  )
  
  all_results <- list()
  
  for (endpoint in names(endpoint_definitions)) {
    time_col  <- endpoint_definitions[[endpoint]]$time
    event_col <- endpoint_definitions[[endpoint]]$event
    
    # Modified dataframe for 5-year truncation
    df_mod <- df %>%
      mutate(
        !!time_col := ifelse(!!sym(time_col) > 60, 60, !!sym(time_col)),
        !!event_col := ifelse(!!sym(time_col) >= 60, 0, !!sym(event_col))
      )
    
    for (five_years in c(FALSE, TRUE)) {
      current_df <- if (five_years) df_mod else df
      
      res <- map_dfr(pathway_cols, function(pathway_col) {
        pathway_name <- names(df)[pathway_col]
        
        formula <- as.formula(paste0("Surv(`", time_col, "`, `", event_col, "`) ~ `", pathway_name, "`"))
        
        cox_model <- tryCatch(
          suppressWarnings(coxph(formula, data = current_df)),
          error = function(e) NULL
        )
        
        if (!is.null(cox_model)) {
          coef_summary <- summary(cox_model)$coefficients
          tibble(
            pathway = pathway_name,
            HR = coef_summary[1, "exp(coef)"],
            p_value = coef_summary[1, "Pr(>|z|)"]
          )
        } else {
          tibble(
            pathway = pathway_name,
            HR = NA,
            p_value = NA
          )
        }
      })
      
      # Adjust p-values for FDR
      res <- res %>%
        mutate(FDR = p.adjust(p_value, method = "fdr"),
               endpoint = if (five_years) paste0(endpoint, "-5year") else endpoint)
      
      all_results[[if (five_years) paste0(endpoint, "-5year") else endpoint]] <- res
    }
  }
  
  # Save results into one Excel file if requested
  if (save_results) {
    wb <- createWorkbook()
    
    for (endpoint_name in names(all_results)) {
      addWorksheet(wb, sheetName = endpoint_name)
      writeData(wb, sheet = endpoint_name, all_results[[endpoint_name]])
    }
    
    file_name <- paste0(output_dir, "/", "cox_results_", if (!is.null(tag)) paste0(tag, "_"), "PVal_endpoints.xlsx")
    saveWorkbook(wb, file = file_name, overwrite = TRUE)
  }
  
  final_results <- bind_rows(all_results)
  
  return(final_results)
}

```
# Section 1-2: Employ the HR function on you data
```{r}
disc_erpos <- read.delim("DISC.ERpos.txt", 
                         check.names = FALSE,
                         row.names = 1)
disc_erpos_hr <- analyze_pathways_survival(
  df = disc_erpos,
  pathway_cols = 10:99,
  tag = "DISC_ERpos",
  save_results = TRUE,
  output_dir = "."
)

disc.erneg <- read.delim("DISC.ERneg.txt", 
                         check.names = FALSE,
                         row.names = 1)
disc_erneg_hr <- analyze_pathways_survival(
  df = disc.erneg,
  pathway_cols = 10:99,
  tag = "DISC_ERneg",
  save_results = TRUE,
  output_dir = "."
)
valid.erpos <- read.delim("VALID.ERpos.txt", 
                         check.names = FALSE,
                         row.names = 1)
valid_erpos_hr <- analyze_pathways_survival(
  df = valid.erpos,
  pathway_cols = 10:99,
  tag = "VALID_ERpos",
  save_results = TRUE,
  output_dir = "."
)
valid.erneg <- read.delim("VALID.ERneg.txt", 
                         check.names = FALSE,
                         row.names = 1)
valid_erneg_hr <- analyze_pathways_survival(
  df = valid.erneg,
  pathway_cols = 10:99,
  tag = "VALID_ERneg",
  save_results = TRUE,
  output_dir = "."
)

```
# Section 2: Helper function to Find common significant pathways
```{r}
find_common_significant_pathways <- function(disc_data, valid_data) {
  
  # Filter disc and valid for p_value <0.05
  disc_sig <- disc_data %>%
    filter(p_value <0.05)
  
  valid_sig <- valid_data %>%
    filter(p_value <0.05)
  
  # Inner join by pathway and endpoint
  common <- inner_join(
    disc_sig,
    valid_sig,
    by = c("pathway", "endpoint"),
    suffix = c("_disc", "_valid")
  )
  
  # Keep pathways where HR is consistently >1 or consistently <1
  common_consistent <- common %>%
    filter(
      (HR_disc > 1 & HR_valid > 1) |
        (HR_disc < 1 & HR_valid < 1)
    )
  
  return(common_consistent)
}
```
# Section 2-1: Employ the common significant pathways function
```{r}
er_pos_common <- find_common_significant_pathways(disc_erpos_hr, valid_erpos_hr)
er_neg_common <- find_common_significant_pathways(disc_erneg_hr, valid_erneg_hr)
```

# Section 3: Prepare data for plotting
```{r}
# Add ER status
er_pos_common <- er_pos_common %>% mutate(ER_Status = "ER-Positive")
er_neg_common <- er_neg_common %>% mutate(ER_Status = "ER-Negative")

# Combine both datasets
common_results_combined <- bind_rows(er_pos_common, er_neg_common)

# Identify common (pathway, endpoint) pairs
shared_pairs <- inner_join(
  er_pos_common %>% select(pathway, endpoint) %>% distinct(),
  er_neg_common %>% select(pathway, endpoint) %>% distinct(),
  by = c("pathway", "endpoint")
)

# Mark shared pathway-endpoint pairs as "Common"
common_results_combined <- common_results_combined %>%
  mutate(ER_Status = ifelse(paste(pathway, endpoint) %in% 
                              paste(shared_pairs$pathway, shared_pairs$endpoint),
                            "Common", ER_Status))

# Select p-value/HR
common_results_combined <- common_results_combined %>%
  mutate(
    selected_p_value = pmax(p_value_disc, p_value_valid, na.rm = TRUE),
    selected_HR = ifelse(p_value_disc >= p_value_valid, HR_disc, HR_valid),
    selected_dataset = ifelse(p_value_disc >= p_value_valid, "Discovery", "Validation")
  )

# Save output
write.table(common_results_combined, file = paste0("Common_Pathways_", Sys.Date(), ".txt"), 
            sep = "\t", row.names = FALSE, quote = FALSE)

# Factor levels for pathways
all_pathways <- sort(unique(common_results_combined$pathway))
common_results_combined$pathway <- factor(common_results_combined$pathway, levels = all_pathways)

er_colors <- c("Common" = "mediumorchid4", "ER-Positive" = "steelblue1", "ER-Negative" = "indianred3")

# Bubble sizes
common_results_combined <- common_results_combined %>%
  mutate(SizeMetric = -log10(selected_p_value),
         SizeMetric2 = SizeMetric * 1.5)

# Grey square condition
common_results_combined <- common_results_combined %>%
  mutate(Grey_Square = (HR_disc < 1 & HR_valid < 1))

# Endpoint order
endpoint_order <- c("MFS-5year", "RFS-5year", "DSS-5year", "OS-5year",
                    "MFS", "RFS", "DSS", "OS")
common_results_combined$endpoint <- factor(common_results_combined$endpoint, levels = endpoint_order)

```

# ⚠️ if you ever change endpoint names, you must manually edit it
# section 4-1: Create plots for all endpoints
```{r}

p_all <- ggplot(common_results_combined, aes(x = pathway, y = endpoint)) +
    geom_point(aes(size = SizeMetric, color = ER_Status), shape = 16, fill = NA) +
    scale_color_manual(values = er_colors,
                       breaks = c("Common","ER-Positive","ER-Negative")) +
    scale_size_continuous(name = "-log (P-value)", range = c(2, 8)) +
    scale_x_discrete(expand = expansion(add = c(0.5, 0.7))) + 
    labs(x = "Pathway", y = "Survival Type", color = "ER Status") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(size = 12, angle = 90, hjust = 1, vjust = 0.15),
      axis.title.x = element_text(hjust = 0.5, vjust = -3.5, size = 14),
      axis.title.y = element_text(hjust = 0.5, vjust = 5, angle = 90, size = 14),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 30),
      panel.border = element_rect(color = "grey", fill = NA, linewidth = 1)
    ) +
    guides(color = guide_legend(order = 1), size = guide_legend(order = 2))

  # Add grey squares, subset from correct `data`
  p_all <- p_all + geom_point(data = subset(common_results_combined, Grey_Square),
                      aes(x = pathway, y = endpoint, size = SizeMetric2),
                      shape = 22, fill = NA, color = "grey", stroke = 2,
                      show.legend = FALSE)

# Custom grey square legend
  legend_plot_all <- ggdraw() + 
    draw_label("HR < 1", x = 0.95, y = 6.75, size = 10, fontface = "italic") +
    draw_grob(rectGrob(gp = gpar(col = "grey", fill = NA, lwd = 3)),
              x = 0.913, y = 6.6, width = 0.015, height = 0.3)

  # Combine plots
  final_plot_all <- plot_grid(p_all, legend_plot_all, ncol = 1, rel_heights = c(1, 0.1))

  # Show plot
  print(final_plot_all)

  # Save the plot with correct filename
  ggsave(filename = paste0("All", "_Survival_Bubble_", Sys.Date(), ".TIFF"),
         plot = final_plot_all, width = 14, height = 8)

```

# Section 4-2:  Create plots for overall survival  
```{r}
overall_data <- common_results_combined %>%
  filter(!grepl("-5year$", endpoint))

p_overall <- ggplot(overall_data, aes(x = pathway, y = endpoint)) +
    geom_point(aes(size = SizeMetric, color = ER_Status), shape = 16, fill = NA) +
    scale_color_manual(values = er_colors,
                       breaks = c("Common","ER-Positive","ER-Negative")) +
    scale_size_continuous(name = "-log (P-value)", range = c(2, 8)) +
    scale_x_discrete(expand = expansion(add = c(0.5, 0.7))) + 
    scale_y_discrete(expand = expansion(mult = c(0.2,0.2))) +
    labs(x = "Pathway", y = "Survival Type", color = "ER Status") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(size = 12, angle = 90, hjust = 1, vjust = 0.15),
      axis.title.x = element_text(hjust = 0.5, vjust = -3.5, size = 14),
      axis.title.y = element_text(hjust = 0.5, vjust = 5, angle = 90, size = 14),
      plot.margin = margin(t = 30, r = 10, b = 10, l = 30),
      panel.border = element_rect(color = "grey", fill = NA, linewidth = 1)
    ) +
    guides(color = guide_legend(order = 1), size = guide_legend(order = 2))

  # Add grey squares, subset from correct `data`
  p_overall <- p_overall +
    geom_point(data = subset(overall_data, Grey_Square),
                      aes(x = pathway, y = endpoint, size = SizeMetric2),
                      shape = 22, fill = NA, color = "grey", stroke = 2,
                      show.legend = FALSE)

# Custom grey square legend
  legend_plot_overall <- ggdraw() + 
    draw_label("HR < 1", x = 0.95, y = 57, size = 10, fontface = "italic") +
    draw_grob(rectGrob(gp = gpar(col = "grey", fill = NA, lwd = 3)),
              x = 0.913, y = 55.5, width = 0.013, height = 2.5)

  # Combine plots
  final_plot_overall <- plot_grid(p_overall, legend_plot_overall, ncol = 1, rel_heights = c(1, 0.01))

  # Show plot
  print(final_plot_overall)

  # Save the plot with correct filename
  ggsave(filename = paste0("Just_overall", "_Survival_Bubble_", Sys.Date(), ".TIFF"),
         plot = final_plot_overall, width = 14, height = 7)

```

# section 4-3 : Create 5-year survival plots
```{r}
five_year_data <- common_results_combined %>%
  filter(grepl("-5year$", endpoint))

p_5year <- ggplot(five_year_data, aes(x = pathway, y = endpoint)) +
    geom_point(aes(size = SizeMetric, color = ER_Status), shape = 16, fill = NA) +
    scale_color_manual(values = er_colors,
                       breaks = c("Common","ER-Positive","ER-Negative")) +
    scale_size_continuous(name = "-log (P-value)", range = c(2, 8)) +
  #space between columns :  
  scale_x_discrete(expand = expansion(add = c(0.5, 0.7))) + 
  #space between rows : or expansion(add = c(2, 2))
    scale_y_discrete(expand = expansion(mult = c(0.15,0.15))) +
    labs(x = "Pathway", y = "Survival Type", color = "ER Status") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(size = 12, angle = 90, hjust = 1, vjust = 0.15),
      axis.title.x = element_text(hjust = 0.5, vjust = -3.5, size = 14),
      axis.title.y = element_text(hjust = 0.5, vjust = 5, angle = 90, size = 14),
      #Box size
      plot.margin = margin(t = 25, r = 1, b = 10, l = 20),
      panel.border = element_rect(color = "grey", fill = NA, linewidth = 1)
    ) +
    guides(color = guide_legend(order = 1), size = guide_legend(order = 2))

  # Add grey squares, subset from correct `data`
  p_5year <- p_5year + geom_point(data =  subset(five_year_data, Grey_Square),
                      aes(x = pathway, y = endpoint, size = SizeMetric2),
                      shape = 22, fill = NA, color = "grey", stroke = 2,
                      show.legend = FALSE)

# Custom grey square legend
  legend_plot_5year <- ggdraw() + 
    draw_label("HR < 1",  x = 0.957, y = 48.5,  size = 10, fontface = "italic") +
    draw_grob(rectGrob(gp = gpar(col = "grey", fill = NA, lwd = 3)),
              x = 0.923, y = 47, width = 0.0115, height = 3)

  # Combine plots
  final_plot_5year <- plot_grid(p_5year, legend_plot_5year, ncol = 1, rel_heights = c(1, 0.01))

  # Show plot
  print(final_plot_5year)

  # Save the plot with correct filename
  ggsave(filename = paste0("Just_5year", "_Survival_Bubble_", Sys.Date(), ".TIFF"),
         plot = final_plot_5year, width = 14, height = 5)


```


<details>
<summary>⚙️session info⚙️</summary>

# sessionInfo
```{r}
sessionInfo()
```

# Save everything
```{r}
# Save R session
save(list = ls(), file = paste0("Final_HR_ER_", Sys.Date(), "_PVal.RData"))
```

